---
title: "Illumina BeadChip processing: TXT + BGX → gene-level (median)"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

This notebook ingests **multiple Illumina BeadChip sample files** (e.g., `GSM*.txt.gz`) and a **platform BGX** file (e.g., `GPL6947_*.bgx.gz`), then produces:
- A merged **probe-level** matrix from `AVG_Signal` columns (one column per sample)
- Optional **detection p-value** matrix
- **Gene-level** median summary using platform annotation from the BGX
- CSV outputs in an `output_illumina_bgx` folder

> Assumptions:
> - Files are in one directory (set `data_dir` below).
> - Sample files contain columns like `AVG_Signal` and `Detection Pval`.
> - Probe key is **Array_Address_Id** (preferred) or **ProbeID** (fallback).
> - Platform annotation is provided via a **BGX** file (e.g., GPL6947).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)
```

## 1) Packages

```{r packages}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
pkgs <- c("illuminaio","dplyr","readr","stringr","purrr","tidyr","Biobase")
for (p in pkgs) {
  if (!suppressWarnings(require(p, character.only = TRUE))) {
    BiocManager::install(p, ask = FALSE, update = FALSE)
    library(p, character.only = TRUE)
  }
}
```

## 2) Paths: set your data directory

```{r paths}
# CHANGE this to your folder containing BGX + GSM*.txt.gz files
data_dir <- "GSE30101_RAW/CEL_files"

# Auto-detect BGX (gz or plain)
bgx_file <- list.files(data_dir, pattern = "\\.bgx(\\.gz)?$", full.names = TRUE)
if (length(bgx_file) == 0) stop("No BGX file found in data_dir.")
bgx_file <- bgx_file[1]  # if multiple, take the first for now

# Sample files (gz or plain)
sample_files <- list.files(data_dir, pattern = "^GSM.*\\.txt(\\.gz)?$", full.names = TRUE)
if (length(sample_files) == 0) stop("No GSM*.txt(.gz) files found in data_dir.")
length(sample_files); head(sample_files)
```

## 3) Read platform annotation (BGX)

```{r read_bgx}
bgx <- illuminaio::readBGX(bgx_file)

# Inspect columns
names(bgx)
names(bgx$Probes)

# Heuristic selection of key & symbol columns in BGX
probe_df <- bgx$probes

# Identify the probe key column:
key_candidates <- c("ArrayAddressID","Array_Address_Id","ProbeID","Probe_Id","PROBE_ID","ProbeId")
key_col <- key_candidates[key_candidates %in% names(probe_df)]
if (length(key_col) == 0) stop("Could not find probe key column in BGX Probes table.")
key_col <- key_col[1]

# Identify a gene symbol column:
symbol_candidates <- c("Symbol","ILMN_Gene","GeneSymbol","Gene Symbol","SYMBOL")
sym_col <- symbol_candidates[symbol_candidates %in% names(probe_df)]
if (length(sym_col) == 0) { sym_col <- NA_character_ }

anno <- probe_df[, unique(na.omit(c(key_col, sym_col))) , drop = FALSE]
colnames(anno)[1] <- "ProbeKey"
if (ncol(anno) > 1) colnames(anno)[2] <- "SYMBOL" else anno$SYMBOL <- NA_character_

# Drop duplicates in annotation
anno <- anno[!duplicated(anno$ProbeKey), ]
head(anno)
```

## 4) Read and merge all sample TXT files

```{r read_merge}
read_one_sample <- function(fpath) {
  df <- suppressMessages(readr::read_tsv(fpath, comment = "#", guess_max = 200000))
  cn <- names(df)
  
  # Probe key column is the first column (...1)
  key_col <- cn[1]
  
  # Identify AVG_Signal and Detection_Pval columns
  avg_col <- cn[grepl("AVG_Signal$", cn)]
  det_col <- cn[grepl("Detection_Pval$", cn)]
  
  if (length(avg_col) == 0) stop(paste("No AVG_Signal column found in:", fpath))
  avg_col <- avg_col[1]
  det_col <- if (length(det_col) > 0) det_col[1] else NA_character_
  
  # Create a cleaned sample name based on file basename
  sample_name <- gsub("\\.txt(\\.gz)?$", "", basename(fpath))
  
  # Extract expression
  out <- df[, c(key_col, avg_col), drop = FALSE]
  colnames(out) <- c("ProbeKey", sample_name)
  
  # Extract detection p-values (if available)
  if (!is.na(det_col)) {
    out_det <- df[, c(key_col, det_col), drop = FALSE]
    colnames(out_det) <- c("ProbeKey", paste0(sample_name, ".detP"))
  } else {
    out_det <- NULL
  }
  
  list(signal = out, detp = out_det)
}


lst <- lapply(sample_files, read_one_sample)
signal_list <- lapply(lst, `[[`, "signal")
detp_list   <- lapply(lst, `[[`, "detp")

signal_merged <- Reduce(function(x,y) dplyr::full_join(x,y, by = "ProbeKey"), signal_list)

detp_merged <- NULL
if (all(!vapply(detp_list, is.null, logical(1)))) {
  detp_merged <- Reduce(function(x,y) dplyr::full_join(x,y, by = "ProbeKey"), detp_list)
}

signal_merged <- signal_merged[!duplicated(signal_merged$ProbeKey), ]
rownames(signal_merged) <- signal_merged$ProbeKey

if (!is.null(detp_merged)) {
  detp_merged <- detp_merged[match(rownames(signal_merged), detp_merged$ProbeKey), ]
  rownames(detp_merged) <- detp_merged$ProbeKey
}

dim(signal_merged); head(signal_merged[, 1:5])
if (!is.null(detp_merged)) { dim(detp_merged); head(detp_merged[, 1:5]) }
```
```{r}
df_test <- readr::read_tsv("GSE30101_RAW/CEL_files/GSM744835.txt.gz", n_max = 10)
colnames(df_test)

```

## 5) Preprocess: log2 + optional detection filtering

```{r preprocess}
rownames(signal_merged) = signal_merged$ProbeKey

exprs_mat <- as.matrix(signal_merged[, setdiff(names(signal_merged), "ProbeKey")])
storage.mode(exprs_mat) <- "double"

rownames(exprs_mat) = rownames(signal_merged)
# Log2 transform if clearly on raw scale
if (is.finite(max(exprs_mat, na.rm = TRUE)) && max(exprs_mat, na.rm = TRUE) > 50) {
  exprs_mat <- log2(exprs_mat + 1)
}

# Detection filter (keep probes with p<0.05 in >=50% samples)
filter_fraction <- 0.1
det_thresh <- 0.1
keep_idx <- rep(TRUE, nrow(exprs_mat))
if (!is.null(detp_merged)) {
  det_mat <- as.matrix(detp_merged[, setdiff(names(detp_merged), "ProbeKey")])
  storage.mode(det_mat) <- "double"
  keep_idx <- rowMeans(det_mat < det_thresh, na.rm = TRUE) >= filter_fraction
}

exprs_mat_filt <- exprs_mat[keep_idx, , drop = FALSE]
probe_keys <- rownames(exprs_mat_filt)
dim(exprs_mat_filt)
```

## 6) Annotate via BGX and summarise median per gene

```{r annotate_summarise}
# 1. Clean probe IDs ONCE and keep them as characters
clean_ids <- sub(".*?_", "", rownames(exprs_mat_filt))

# 2. Apply cleaned probe IDs to the expression matrix rownames
rownames(exprs_mat_filt) <- clean_ids

# 3. Also clean IDs in the annotation table to match
anno$ProbeKey_clean <- sub(".*?_", "", anno$ProbeKey)

# 4. Now match annotation safely
anno_sub <- anno[match(clean_ids, anno$ProbeKey_clean), , drop = FALSE]

# 5. Combine expression + annotation without integer conversion
exprs_df <- cbind(ProbeKey = clean_ids, as.data.frame(exprs_mat_filt, check.names = FALSE))

# 6. Annotate
annotated <- dplyr::left_join(exprs_df, anno_sub, by = c("ProbeKey" = "ProbeKey_clean"))

# 7. Summarize at gene level
gene_expr <- annotated %>%
  dplyr::filter(!is.na(SYMBOL) & SYMBOL != "") %>%
  dplyr::group_by(SYMBOL) %>%
  dplyr::summarise(dplyr::across(where(is.numeric), median, na.rm = TRUE), .groups = "drop")

```

## 7) Save outputs

```{r save}
out_dir <- file.path("output_GSE30101")
dir.create(out_dir, showWarnings = FALSE)

readr::write_csv(data.frame(ProbeKey = rownames(exprs_mat), exprs_mat, check.names = FALSE),
                 file.path(out_dir, "probe_level_matrix.csv"))
readr::write_csv(gene_expr, file.path(out_dir, "gene_level_median_matrix.csv"))

if (!is.null(detp_merged)) {
  det_mat <- as.matrix(detp_merged[, setdiff(names(detp_merged), "ProbeKey")])
  readr::write_csv(data.frame(ProbeKey = rownames(det_mat), det_mat, check.names = FALSE),
                   file.path(out_dir, "detection_pvalues_matrix.csv"))
}

list.files(out_dir, full.names = TRUE)
```

## 8) Quick QC

```{r qc, fig.height=4.5, fig.width=6.5}
boxplot(exprs_mat, las = 2, cex.axis = 0.7, main = "Sample distributions (log2)")

pca <- prcomp(t(exprs_mat), center = TRUE, scale. = TRUE)
plot(pca$x[,1], pca$x[,2], pch = 19,
     xlab = paste0("PC1 (", round(100*summary(pca)$importance[2,1],1), "%)"),
     ylab = paste0("PC2 (", round(100*summary(pca)$importance[2,2],1), "%)"),
     main = "PCA on merged expression")
text(pca$x[,1], pca$x[,2], labels = colnames(exprs_mat), pos = 3, cex = 0.7)
```

## 9) Session info

```{r session}
sessionInfo()
```
